Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2015-01-05 parts/def-ph.tex
fixes after comments from Morgan

    \TState{a_1, b_0, c_0}
  \end{tikzpicture}
  }
  \caption{%
  \label{fig:ph-livelock}%
    An example of AAN.
    This model represents the interaction of two exclusive components $a$ and $b$,
    that cannot be active simultaneously, and that degrade over time.
    Furthermore, these two components can cooperate to activate $c$
    if their “active” states ($a_1$ and $b_1$)
    are present in the same state.
    Automata are represented by labelled boxes
    and local states by circles with their identifier on the side.
    Actions are represented by a dot connected by an edge to every hitter
    and by an arrows to the target, followed by another dotted arrow towards the bounce.
    Greyed local states stand for the following possible global state:
    $\PHstate{a_1, b_0, c_0, ab_{10}}$.
  }
\end{figure}

@@

\subsection{Modelling cooperation}
\label{ssec:cooperation}

Cooperations between components are a very common in biological systems.
They represent the fact that a given regulation is not conditioned by only one condition,
such as the presence of a given concentration of one substance,
represented in the model by the activity of the one related local state.
Instead, in the case of a cooperation, two conditions or have to be bet.
For example, in the example of \pref{fig:ph-livelock},
the activation of component $c$ requires the cooperation of $a$ and $b$;
in practice, this is expressed by a set of two hitters for action
$\PHfrappes{a_1, b_1}{c_0}{c_1}$.

Such a cooperation can be represented in Process Hitting by using a \emph{cooperative sort},
whose aim is to represent a Boolean formula~\cite{PMR10-TCSB}.
The different local states of a cooperative sort represent the possible configurations
of a set of automata,
and actions allow to update the active local state of the cooperative sort
in order to have it match the configuration of these automata.
Then, one or several actions can be added from the desired local states of the cooperative sort
in order to abstract multiple requirements.
Cooperative sorts have the advantage to not require any supplementary semantics
than those of standard Process Hitting,
but at the cost of slightly over-approximating the expected dynamics of a true Boolean gate.
Indeed, nothing forces these actions to be played as soon as possible,
thus resulting in a possible temporal shift between the evolution
of the active process of the cooperative sorts and those of the automata it represents.
In other words, a cooperative sorts does not necessarily represent the configuration
of a given set of automata, but only a combination of past states.

In~\cite{FPMR13-CS2Bio}, the formalization of classes of priorities into the Process Hitting
was proposed in order to add some missing features,
such as the possibility to abstract time properties of regulation or reaction rates.
Another consequence of the addition of priorities was the possibility to fix
the aforementioned temporal shift, and thus be able to model accurate Boolean gates,
by prioritizing the actions used to update the cooperative sorts.
The AAN formalism proposed in this paper have in fact the same expressivity than
Process Hitting with classes of priorities.
Indeed, the actions with multiple hitters play the role of cooperative sorts,
and the classes of priorities can be encoded into AAN without priorities,
as detailed in \pref{ssec:flattening}.

We note that obviously, Process Hitting~\cite{PMR10-TCSB} being a restriction of AANs,
one can easily encode any Process Hitting model as an AAN by using only actions
$\PHfrappe{A}{b_j}{b_k}$ so that $0 \leq \card{A} \leq 1$.
Process Hitting with classes of priorities~\cite{FPMR13-CS2Bio} can also be encoded as AANs
with priorities, as defined in \pref{sec:encodings},
by using the same priorities for each action,
and possibly translate them into simple AANs.

>>>>

    \TState{a_1, b_0, c_0}
  \end{tikzpicture}
  }
  \caption{%
  \label{fig:ph-livelock}%
    An example of AAN.
    This model represents the interaction of two exclusive components $a$ and $b$,
    that cannot be active simultaneously, and that degrade over time.
    Moreover, these two components can cooperate to activate $c$
    if their “active” states ($a_1$ and $b_1$)
    are present in the same state.
    Automata are represented by labelled boxes
    and local states by circles with their identifier on the side.
    Actions are represented by a dot connected by an edge to every hitter
    and by an arrows to the target, followed by another dotted arrow towards the bounce.
    Greyed local states stand for the following possible global state:
    $\PHstate{a_1, b_0, c_0, ab_{10}}$.
  }
\end{figure}

@@

\subsection{Comparison with the Process Hitting and Modelling of Cooperations}
\label{ssec:cooperation}

The AAN formalism can be seen as an extension of the Process Hitting framework
previously introduced in~\cite{PMR10-TCSB}.
Indeed, the representation of Process Hitting models is also based on automata
(called “sorts”), local states (called “processes”) and actions.
The particularity of Process Hitting lies in the following restriction:
each action can have at most one hitter;
in other words, a Process Hitting model is an AAN so that
$\forall h \in \PHh, 0 \leq \card{\hitter{h}} \leq 1$.
Conversely, any Process Hitting model can obviously be encoded into an AAN.
% In \pref{ssec:flattening}, we also show that Process Hitting models with
% classes of priorities~\cite{FPMR13-CS2Bio}
% can also be encoded into AANs.

The difference between AAN and Process Hitting especially appears when one
needs to represent cooperations between components.
%are a very common in biological systems.
A cooperations represents the fact that
a given regulation is not conditioned by only one condition,
such as the presence of a given concentration of one substance,
represented in the model by the activity of the one related local state.
Instead, in the case of a cooperation, two conditions or more have to be bet.
For example, in the example of \pref{fig:ph-livelock},
the activation of component $c$ requires the cooperation of $a$ and $b$;
in practice, this is naturally expressed by a set of two hitters for action
$\PHfrappes{a_1, b_1}{c_0}{c_1}$.

Such a cooperation can also be represented in Process Hitting
by using a “cooperative sort”~\cite{PMR10-TCSB},
which is an additional automata
whose aim is to represent a Boolean formula.
The different local states of a cooperative sort represent the possible configurations
of a set of automata,
and actions allow to “update” the active local state of the cooperative sort
in order to have it match the configuration of these automata.
Then, one or several actions can be added from the desired local states of the cooperative sort
in order to abstract multiple requirements.
Cooperative sorts have the advantage to not require any supplementary semantics
than those of standard Process Hitting
(that is, not more that one hitter per action),
but at the cost of slightly over-approximating the expected dynamics of a true Boolean gate.
Indeed, nothing forces the updating actions to be played as soon as possible,
thus resulting in a possible temporal shift between the evolution
of the active process of the cooperative sorts and those of the automata it represents.
In other words, a cooperative sorts does not necessarily represent the current configuration
of a given set of automata, but only a combination of their past states.
AAN thus fixes this temporal shift by removing the need for cooperative sorts,
but at the cost of changing the semantics of the formalism.

In~\cite{FPMR13-CS2Bio}, the formalization of classes of priorities into the Process Hitting
was proposed in order to add some missing features,
such as the possibility to abstract time properties of regulation or reaction rates.
Another consequence of the addition of priorities was the possibility to fix
the aforementioned temporal shift, and thus be able to model accurate Boolean gates,
by prioritizing the actions used to update the cooperative sorts.
The AAN formalism proposed in this paper has in fact the same expressivity than
Process Hitting with classes of priorities.
Indeed, the actions with multiple hitters play the role of cooperative sorts,
and the classes of priorities can be encoded into AAN without priorities,
as detailed in \pref{ssec:flattening}.
% 
% We note that obviously, Process Hitting~\cite{PMR10-TCSB} being a restriction of AANs,
% one can easily encode any Process Hitting model as an AAN by using only actions
% $\PHfrappe{A}{b_j}{b_k}$ so that $0 \leq \card{A} \leq 1$.
% Process Hitting with classes of priorities~\cite{FPMR13-CS2Bio} can also be encoded as AANs
% with priorities, as defined in \pref{ssec:flattening},
% by using the same priorities for each action,
% and possibly translate them into simple AANs.

### 2015-01-05 parts/intro.tex
fixes after comments from Morgan

Besides, one further objective of our work is now %to be more accurate in the description of the studied systems.
to improve the accuracy of the description of the studied systems dynamics.
The idea for this is to introduce timing features into models:
we are interested in taking into account some knowledge about the relative length of some phenomena as it is a way to refute some models (or parameters) that are inconsistent with the observed dynamic behaviours.
In this paper, we are dealing with these timing properties through priorities,
that are based on the simple founding idea that actions with higher priority have to be processed before the ones with lower priority.
Furthermore, due to the Process Hitting framework restrictions, multimolecular reactions were previously not immediately available, but one could simulate them with an encoding called “cooperative sort”.
That encoding however introduces extra reactions,
that produce a temporal shift between the presence of the reactants
and the playability of the reaction.
This is where the priorities become useful, if not necessary:
the extra reactions can for example be given “infinite speed” (highest priority) so that they do not affect the behaviour of “normal” (lower priority) reactions, including the multimolecular ones.
Another approach consists in considering another class of models,
that we call Asynchronous Automata Network,
and that allows to naturally model these cooperations
by defining several requisites for a reaction.
Furthermore, such automata networks are still compatible with the notion of priority,
that can also be used to model different reaction rates in the model.

@@

This paper is an extension of~\cite{FPMR13-CS2Bio} where we
introduced a new extension to the semantics of Process Hitting
by partitioning actions into classes of priorities.
One of our objectives was to reach an accurate representation of cooperating components in the model, that was not fulfilled with the initial semantics.
Although in~\cite{FPMR13-CS2Bio} this was achieved with a particular class of high priority,
the work presented in this paper, however, focuses on Asynchronous Automata Networks.
The simple version of this formalism (without classes of priorities)
is defined in \pref{sec:ph}
and consists of an extension of Process Hitting with classes of priorities
that allows to simplify the formalisation with the same benefits.
In \pref{sec:sa},
we develop an efficient under-approximation of the reachability
of several local states of components,
thus allowing to compute efficient static analysis of the dynamics.
This local reachability under-approximation can be also easily extended to study the reachability of a global or partial state.
Furthermore, we provide a new extended under-approximation method to refine this method
in the case of a successive reachability property.

Furthermore,
we define in \pref{sec:encodings} the notion of Asynchronous Automata Networks
with classes of priorities,
that consists of a convenient way to introduce time parameters
into the models, under the form of preemptions.
We also provide in \pref{sec:encodings} a “flattening”
of any Asynchronous Automata Networks with classes of priorities
into a model without classes of priorities,
thus extending the range of applicability of the static analysis.
We also give in the same section a formal encoding of Asynchronous Discrete Networks,
that is, the multivalued version of Asynchronous Boolean Networks.
These models,
which encompass variables and evolution functions with a limited number of discrete values,
are also called “Logical Networks” in~\cite{Thomas95,deJong02}.
This encoding thus allows to efficiently study this kind of widespread models.

@@

\paragraph*{Functions and least fixed point}
If $A$ and $B$ are sets,
$f : A \rightarrow B$ denotes an application $f$ that maps the elements of $A$ to elements of $B$.
Furthermore, if $f$ is a monotonically increasing and bounded function, then
$\lfp{x_0}{x}{x'}$ is the least fixed point of the function $x \mapsto x'$ which is greater than $x_0$.

>>>>

Besides, one further objective of our work is now %to be more accurate in the description of the studied systems.
to improve the accuracy of the description of the studied systems dynamics.
The idea for this is to introduce timing features into models:
we are interested in taking into account some knowledge about the relative length of some phenomena as it is a way to refute some models (or parameters) that are inconsistent with the observed dynamic behaviours.
In this paper, we are dealing with these timing properties through priorities,
that are based on the simple founding idea that actions with higher priority have to be processed before the ones with lower priority.
Furthermore, due to the Process Hitting framework restrictions, multimolecular reactions were previously not immediately available, but one could simulate them with an encoding called “cooperative sort”.
That encoding however introduces extra reactions,
that produce a temporal shift between the presence of the reactants
and the playability of the reaction.
This is where the priorities become useful, if not necessary:
the extra reactions can for example be given “infinite speed” (highest priority) so that they do not affect the behaviour of “normal” (lower priority) reactions, including the multimolecular ones.
Another approach consists in considering another class of models,
that we call Asynchronous Automata Network,
and that allows to naturally model these cooperations
by defining several requisites for a reaction.
Moreover, such automata networks are still compatible with the notion of priority,
that can also be used to model different reaction rates in the model.

@@

This paper is an extension of~\cite{FPMR13-CS2Bio} where we
introduced a new extension to the semantics of Process Hitting
by partitioning actions into classes of priorities.
One of our objectives was to reach an accurate representation of cooperating components in the model, that was not fulfilled with the initial semantics.
Although in~\cite{FPMR13-CS2Bio} this was achieved with a particular class of high priority,
the work presented in this paper focuses instead on Asynchronous Automata Networks.
The simple version of this formalism (without classes of priorities)
is defined in \pref{sec:ph}
and is equivalent to an extension of Process Hitting with classes of priorities,
but allows to simplify the formalisation with the same benefits.
In \pref{sec:sa},
we develop an efficient under-approximation of the reachability
of several local states of components,
thus allowing to compute efficient static analysis of the dynamics.
This local reachability under-approximation can be also easily extended to study the reachability of a global or partial state.
Furthermore, we provide a new extended under-approximation method to refine this method
in the case of a successive reachability property.

Moreover,
we define in \pref{sec:encodings} the notion of Asynchronous Automata Networks
with classes of priorities,
that consists of a convenient way to introduce time parameters
into the models, under the form of preemptions.
We also provide in \pref{sec:encodings} a “flattening”
of any Asynchronous Automata Networks with classes of priorities
into a model without classes of priorities,
thus extending the range of applicability of the static analysis.
We also give in the same section a formal encoding of Asynchronous Discrete Networks,
that is, the multivalued version of Asynchronous Boolean Networks.
These models,
which encompass variables and evolution functions with a limited number of discrete values,
are also called “Logical Networks” in~\cite{Thomas95,deJong02}.
This encoding thus allows to efficiently study this kind of widespread models.

@@

\paragraph*{Functions and least fixed point}
If $A$ and $B$ are sets,
$f : A \rightarrow B$ denotes a function $f$ that maps the elements of $A$ to elements of $B$.
If $f$ is a monotonically increasing and bounded function, then
$\lfp{x_0}{x}{x'}$ is the least fixed point of the function $x \mapsto x'$ which is greater than $x_0$.

### 2015-01-05 parts/models-ph-php.tex
fixes after comments from Morgan

Because we only use classical logic operators, the formulas of Boolean logic on 
distributivity, associativity and commutativity can be used, together with De Morgan's laws on negation.
We also have the following property for the negation of an atom:
\[\forall a \in \ov{\PHs}, \forall a_i \in \ov{\PHl}_a, \forall \mysigma \in \PHsubl,
  \Fsem{\neg a_i}{\mysigma} \Leftrightarrow \Fsem{\bigvee_{\substack{a_j \in \ov{\PHl}_a\\a_j \neq a_i}} a_j}{\mysigma}\]
Indeed, if a local state is not active in a state, this means that another local state of the same automaton is active.
Furthermore, we note that:
\[\forall a \in \ov{\PHs}, \forall a_i, a_j \in \ov{\PHl}_a,
  a_i \neq a_j \Longrightarrow a_i \wedge a_j \equiv \bot \]
because two different processes can never be active simultaneously.

@@

We showed in this subsection that it is possible to model any AAN$k$
as an AAN (or, equivalently, as an AAN$1$).
this translation thus extends the applicability of the static analysis developed in
\pref{sec:sa} to any AAN$k$, with $k \in \sN^*$.
The translation given in this section
is exponential in the number of actions of higher priority for each action.

>>>>

Because we only use classical logic operators, the formulas of Boolean logic on 
distributivity, associativity and commutativity can be used, together with De Morgan's laws on negation.
We also have the following property for the negation of an atom:
\[\forall a \in \ov{\PHs}, \forall a_i \in \ov{\PHl}_a, \forall \mysigma \in \PHsubl,
  \Fsem{\neg a_i}{\mysigma} \Leftrightarrow \Fsem{\bigvee_{\substack{a_j \in \ov{\PHl}_a\\a_j \neq a_i}} a_j}{\mysigma}\]
Indeed, if a local state is not active in a state, this means that another local state of the same automaton is active.
Moreover, we note that:
\[\forall a \in \ov{\PHs}, \forall a_i, a_j \in \ov{\PHl}_a,
  a_i \neq a_j \Longrightarrow a_i \wedge a_j \equiv \bot \]
because two different processes can never be active simultaneously.

@@

We showed in this subsection that it is possible to model any AAN$k$
as an AAN (or, equivalently, as an AAN$1$).
This translation thus extends the applicability of the static analysis developed in
\pref{sec:sa} to any AAN$k$, with $k \in \sN^*$.
Moreover, it allows to represent
any Process Hitting model with classes of priorities~\cite{FPMR13-CS2Bio}
under the form of an AAN
(or, equivalently, of a Process Hitting model with $2$ classes of priorities).
The translation given in this section
is exponential in the number of actions of higher priority for each action.
